Data exploration

### 2021-07-01 ms/data_exploration.Rmd
Only use dataset for data exploration

There are two data sets:
* `biodiv_ferns_cent_env` with biodiversity metrics measured on the full set of species (species missing functional data were excluded when measuring functional diversity metrics).
* `biodiv_ferns_cent_repro` with biodiveristy metrics related to genetic diversity measured only on species with reproductive mode data available.

@@

Below, "Full dataset" refers to `biodiv_ferns_cent_env` and "Repro dataset" refers to `biodiv_ferns_cent_repro`.

```{r load-data}
# Load the data from the targets workflow
tar_load(biodiv_ferns_cent_env)
tar_load(biodiv_ferns_cent_repro)
```

@@

```{r struc-full-data}
biodiv_ferns_cent_env
```

```{r struc-repro-data}
biodiv_ferns_cent_repro
```

@@

* `grids`: ID code for each grid cell
* `lat`: latitude of grid cell centroid
* `long`: longitude of grid cell centroid
* `richness`: number of taxa
* `pd_obs_z`: standard effect size (SES) of phylogenetic diversity (pd)
* `fd_obs_z`: SES of functional diversity (fd)
* `rpd_obs_z`: SES of relative pd
* `rfd_obs_z`: SES of relative fd
* `temp`: mean annual temperature (10 x °C)
* `precip`: total annual precipitation (mm)
* `precip_season`: seasonality in precipitation
* `percent_apo`: percent of apogamous taxa in that grid cell (repro data only)

@@

```{r check-missing}
# These will return TRUE if all data are non-missing, 
# or an error if any data are missing (NA)
biodiv_ferns_cent_env %>%
  assert(not_na, everything(), success_fun = success_logical)

biodiv_ferns_cent_repro %>%
  assert(not_na, everything(), success_fun = success_logical)
```

@@

```{r box-plots-full}
biodiv_ferns_cent_env %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  facet_wrap(vars(variable), scales = "free") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  ) +
  labs(title = "Full dataset")
```

```{r box-plots-repro}
biodiv_ferns_cent_repro %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  facet_wrap(vars(variable), scales = "free") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  ) +
  labs(title = "Repro dataset")
```

@@

```{r grubbs-env}
biodiv_ferns_cent_env %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  group_by(variable) %>%
  nest() %>%
  mutate(map_df(data, ~grubbs_tidy(.$value))) %>%
  select(-data) %>%
  ungroup() %>%
  select(variable:p_value) %>%
  mutate(signif = p_value < 0.05)
```

```{r grubbs-repro}
biodiv_ferns_cent_repro %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  group_by(variable) %>%
  nest() %>%
  mutate(map_df(data, ~grubbs_tidy(.$value))) %>%
  select(-data) %>%
  ungroup() %>%
  select(variable:p_value) %>%
  mutate(signif = p_value < 0.05)
```

@@

```{r check-apo-outlier}
biodiv_ferns_cent_repro %>%
  slice_max(1, order_by = percent_apo) %>%
  left_join(select(biodiv_ferns_cent_env, grids, richness)) %>%
  select(grids:long, percent_apo, richness)
```

@@

```{r remove-apo-outlier-check-grubbs}
biodiv_ferns_cent_repro %>% 
  filter(percent_apo != max(percent_apo)) %>%
  pull(percent_apo) %>%
  grubbs_tidy()
```

@@

```{r hist-env}
biodiv_ferns_cent_env %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(vars(variable), scales = "free") +
  labs(title = "Full dataset")
```

```{r hist-repro}
biodiv_ferns_cent_repro %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(vars(variable), scales = "free") +
  labs(title = "Repro dataset")
```

@@

```{r qqplot-env}
biodiv_ferns_cent_env %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(sample = value)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Full dataset")
```

@@

```{r qqplot-repro}
biodiv_ferns_cent_repro %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(sample = value)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Repro dataset")
```

Again, the response variables `pd_obs_z`, and `rpd_obs_z` look OK.

@@

```{r check-for-zeros}
biodiv_ferns_cent_env %>%
  mutate(rich_zero = richness == 0) %>%
  arrange(richness) %>%
  summarize(
    total_zero = sum(rich_zero),
    perc_zero = sum(rich_zero) / nrow(.))
```

@@

```{r plot-raw-data-env, fig.height = 8}
ggplot(biodiv_ferns_cent_env, aes(x = .panel_x, y = .panel_y)) + 
  geom_point(alpha = 0.5, shape = 16, size = 0.5) + 
  geom_smooth() +
  facet_matrix(
    cols = vars(temp, precip, precip_season),
    rows = vars(richness, pd_obs_z, fd_obs_z, rpd_obs_z, rfd_obs_z)) +
  labs(title = "Full dataset")
```

@@

```{r plot-raw-data-repro}
# Remove the outlier first
biodiv_ferns_cent_repro %>% 
  filter(percent_apo != max(percent_apo)) %>%
ggplot(aes(x = .panel_x, y = .panel_y)) + 
  geom_point(alpha = 0.5, shape = 16, size = 0.5) + 
  geom_smooth() +
  facet_matrix(
    cols = vars(percent_apo, temp, precip, precip_season),
    rows = vars(pd_obs_z, rpd_obs_z)) +
  labs(title = "Repro dataset")
```

@@

```{r check-overdispersion}
# Create general linear model with poisson distribution
richness_poisson_mod <- glm(
  richness ~ temp + I(temp^2) + precip + precip_season, 
  data = biodiv_ferns_cent_env, family = poisson)

>>>>

The dataset includes biodiversity metrics and environmental data.

@@

```{r load-data}
# Load the data from the targets workflow
tar_load(biodiv_ferns_spatial)
tar_load(resp_vars_env)
tar_load(indep_vars)
tar_load(lat_area_ja)

# Make biodiversity metrics dataframe with centroid of each site for models
  biodiv_ferns_cent <- sf_to_centroids(biodiv_ferns_spatial) %>%
    # add area
    add_roll_area(lat_area_ja) %>%
    # keep only variables needed for model and only rows with zero missing data
    filter_data_for_model(c("grids", "lat", "long", resp_vars_env, indep_vars))
```

@@

```{r struc-full-data}
biodiv_ferns_cent
```

@@

* `grids`: ID code for each grid cell
* `lat`: latitude of grid cell centroid
* `long`: longitude of grid cell centroid
* `richness`: number of taxa
* `pd_obs_z`: standard effect size (SES) of phylogenetic diversity (pd)
* `fd_obs_z`: SES of functional diversity (fd)
* `rpd_obs_z`: SES of relative pd
* `rfd_obs_z`: SES of relative fd
* `percent_apo`: percent of apogamous taxa in that grid cell
* `temp`: mean annual temperature (10 x °C)
* `precip`: total annual precipitation (mm)
* `precip_season`: seasonality in precipitation
* `area`: rolling mean of area in 0.2 degree latitudinal bands

@@

```{r check-missing}
# These will return TRUE if all data are non-missing, 
# or an error if any data are missing (NA)
biodiv_ferns_cent %>%
  assert(not_na, everything(), success_fun = success_logical)
```

@@

```{r box-plots-full}
biodiv_ferns_cent %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  facet_wrap(vars(variable), scales = "free") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
```

@@

```{r grubbs-env}
biodiv_ferns_cent %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  group_by(variable) %>%
  nest() %>%
  mutate(map_df(data, ~grubbs_tidy(.$value))) %>%
  select(-data) %>%
  ungroup() %>%
  select(variable:p_value) %>%
  mutate(signif = p_value < 0.05)
```

@@

```{r check-apo-outlier}
biodiv_ferns_cent %>%
  slice_max(1, order_by = percent_apo) %>%
  left_join(select(biodiv_ferns_cent, grids, richness)) %>%
  select(grids:long, percent_apo, richness)
```

@@

```{r remove-apo-outlier-check-grubbs}
biodiv_ferns_cent %>% 
  filter(percent_apo != max(percent_apo)) %>%
  pull(percent_apo) %>%
  grubbs_tidy()
```

@@

```{r hist-env}
biodiv_ferns_cent %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(vars(variable), scales = "free")
```

@@

```{r qqplot-env}
biodiv_ferns_cent %>%
  select(-lat, -long) %>%
  pivot_longer(-grids, names_to = "variable") %>%
  ggplot(aes(sample = value)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Full dataset")
```

@@

@@

```{r check-for-zeros}
biodiv_ferns_cent %>%
  mutate(rich_zero = richness == 0) %>%
  arrange(richness) %>%
  summarize(
    total_zero = sum(rich_zero),
    perc_zero = sum(rich_zero) / nrow(.))
```

@@

```{r plot-raw-data, fig.height = 8}
ggplot(biodiv_ferns_cent, aes(x = .panel_x, y = .panel_y)) + 
  geom_point(alpha = 0.5, shape = 16, size = 0.5) + 
  geom_smooth() +
  facet_matrix(
    cols = vars(temp, precip, precip_season, percent_apo),
    rows = vars(richness, pd_obs_z, fd_obs_z, rpd_obs_z, rfd_obs_z))
```

@@

```{r plot-raw-data-repro}
# Remove the outlier first
biodiv_ferns_cent %>% 
  filter(percent_apo != max(percent_apo)) %>%
ggplot(aes(x = .panel_x, y = .panel_y)) + 
  geom_point(alpha = 0.5, shape = 16, size = 0.5) + 
  geom_smooth() +
  facet_matrix(
    cols = vars(percent_apo, temp, precip, precip_season),
    rows = vars(pd_obs_z, rpd_obs_z))
```

@@

```{r check-overdispersion}
# Create general linear model with poisson distribution
richness_poisson_mod <- glm(
  richness ~ temp + I(temp^2) + precip + precip_season, 
  data = biodiv_ferns_cent, family = poisson)
